---
title: "1.0-Data Cleaning"
output: html_document
---

# Packages

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(labelled)
library(arrow)
library(data.table)
```

# Load the data
This data set stores clinical information from patients in the PREDIMAR study. The aim of this project is to build a machine learning model capable of predicting AF recurrence based on medical data.

Let's load the data set:

```{r echo = FALSE}
global <- read_dta("data/raw/predimar_miguelgomez.dta")
```

Since it is a STATA file, the labels need to be removed:

```{r}
original_clinical_data <- global |> remove_labels()

# Save
write_parquet(
    original_clinical_data, 
    "data/processed/clinical_data_original.parquet"
    )
```

# Data cleaning
The data set consists of thousands of features referring over 700 patients. The ones taken into account are clinical and nutritional characteristics:

* **Clinical data**
** `tipofa`: type of Atrial Fibrillation ($0$ paroxysmal,$1$ recurrent).
** `fec_fa_0m`: date of the first AF diagnosis.
** `fec_ablac_0m`: date of the first catheter ablation.
** `interv`: assigned cohort ($0$ control, $1$ intervention).
** `event18m_conkardia`: whether there has been AF recurrence after the blanking period and before $18$ months since the catheter ablation ($0$ no, $1$ yes).
** `sexo`: gender ($0$ male, $1$ female)
** `edad`: age.
** `ablacion_previa`: whether the patient has undergone catheter ablation previously to this trial ($0$ no, $1$ yes).
** `tiempo_fa_ablac`: years since the first AF diagnosis.
** `dilat_tot4_imp`: left atrial enlargement ($0$ normal, $1$ mild, $2$ moderate, $3$ severe).
** `bmi_autoref_0m`: Body Mass Index ($0$ normal weight, $25$ overweight, $30$ obese).
** `mettotal_0m`: MET (physical activity measurement).
** `hta_0m`: hypertension ($0$ no, $1$ yes).
** `diabetes2_0m`: type 2 diabetes ($0$ no, $1$ yes).
** `hipolipemiantes_0m`: whether the patient takes hypolipidemic medication ($0$ no, $1$ yes).
** `hipercol_0m`: whether the patient suffers from dyslipidemia ($0$ no, $1$ yes).

* **Nutritional data**
** `p14_tot_0m:`: affinity to the Mediterranean diet ($0 - 14$).
** `energiat_0m`: total daily energy intake ($kcal$).
** `ac_oliva_0m`: olive oil daily intake ($g$).
** `ac_olivavir_0m`: extra virgin olive oil daily intake ($g$).
** `ac_orujo_0m`: pomace oil daily intake ($g$).
** `ac_maiz_0m`: corn oil daily intake ($g$).
** `ac_girasol_0m`: sunflower oil daily intake ($g$).
** `ac_soja_0m`: soy oil daily intake ($g$).
** `v_blanco_0m`: white wine daily intake ($g$).
** `v_tintojov_0m`: young red wine daily intake ($g$
** `v_tintoaniej_0m`: aged red wine daily intake ($g$).
** `v_rosado_0m`: rose wine daily intake ($g$).
** `cervezas_0m`: beer daily intake ($g$).
** `alcoholg_0m`: total alcohol daily intake ($g$).

```{r}
clinical_data_selected <- original_clinical_data |> 
    select(

        # Clinical DATA
        code,
        tipofa,
        fec_fa_0m,
        fec_ablac_0m,
        interv,
        nodo,
        sexo,
        edad,
        event18m_conkardia, # PREDICTED CLASS
        ablacion_previa,
        tiempo_fa_ablac,
        hta_0m,
        dilat_tot4_imp,
        bmi_autoref_0m,
        mettotal_0m,
        diabetest2_0m,
        hipolipemiantes_0m,
        hipercol_0m,
        
        # NUTRITIONAL DATA (g/day)
        p14_tot_0m,
        energiat_0m,
        ac_oliva_0m,
        ac_olivavir_0m,
        ac_orujo_0m,
        ac_maiz_0m,
        ac_girasol_0m,
        ac_soja_0m,
        v_blanco_0m,
        v_tintojov_0m,
        v_tintoaniej_0m,
        v_rosado_0m,
        cervezas_0m,
        alcoholg_0m,
    )

head(clinical_data_selected)
```

Lets label the categorical data using factors:

```{r}
clinical_data_modified <- clinical_data_selected |> 
    mutate(
        tipofa = factor(
            tipofa,
            levels = c(0, 1),
            labels = c("paroxysmal", "persistent")
        ),
        interv = factor(
            interv,
            levels = c(0, 1),
            labels = c("control", "intervention")
        ),
        nodo = factor(
            nodo,
            levels = c(1, 2, 3, 4),
            labels = c("Pamplona", "Madrid", "Granada", "Alicante")
        ),
        sexo = factor(
            sexo,
            levels = c(0, 1),
            labels = c("male", "female")     
        ),
        event18m_conkardia = factor(
            event18m_conkardia,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        ablacion_previa = factor(
            ablacion_previa,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        hta_0m = factor(
            hta_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        dilat_tot4_imp = factor(
            dilat_tot4_imp,
            levels = c(0, 1, 2, 3),
            labels = c("normal", "mild", "moderate", "severe")
        ),
        diabetest2_0m = factor(
            diabetest2_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        hipolipemiantes_0m = factor(
            hipolipemiantes_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        hipercol_0m = factor(
            hipercol_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        )
    )

head(clinical_data_modified)
```

The `fec_fa_0m` and `fec_ablac_0m` features can be used to create a new feature called `tiempo_fa_ablac` that measures the years since the first AF diagnosis until the catheter ablation.

```{r}
clinical_data_modified <- clinical_data_modified |> 
    mutate(
        tiempo_fa_ablac = as.numeric(fec_ablac_0m - fec_fa_0m) / 365.25
    ) |> 
    select(-c(fec_fa_0m, fec_ablac_0m))
```

Lets translate and modify the variables' names to make them more understandable:

```{r}
new_names <- c("id", "fa_type", "cohort", "node", "gender", "age", "cardiac_event", "prev_ablation", "fa_ablat_time", "hbp", "atrial_dilation", "bmi", "met", "diabetes_t2", "hypolipidemic", "hypercholest", "p14", "energy", "olive_oil", "ev_olive_oil", "pomace_oil", "corn_oil", "sunflower_oil", "soy_oil", "white_wine", "young_red_wine", "aged_red_wine", "rose_wine", "beer", "alcohol")

clinical_data_clean <- clinical_data_modified |> setnames(
    old = names(clinical_data_modified), 
    new = new_names
    )
```

This is the basal data we will use to build the model. Lets save it:

```{r}
# Save it into the project's data directory
write_parquet(
    clinical_data_clean, 
    "data/processed/clinical_data_clean.parquet"
    )

# Save it into Google drive's data directory
write_parquet(
    clinical_data_clean, 
    "H:/Mi unidad/TFM/data/processed/clinical_data_clean.parquet"
    )

head(clinical_data_clean)
```
