---
title: "1.0-Data processeding"
output: html_document
---

# Packages

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(labelled)
library(arrow)
library(data.table)
```

# Load the data
This data set stores clinical information from patients in the PREDIMAR study. The aim of this project is to build a machine learning model capable of predicting AF recurrence based on medical data.

Let's load the data set:

```{r echo = FALSE}
global <- read_dta("data/raw/predimar_miguelgomez.dta")
```

Since it is a STATA file, the labels need to be removed:

```{r}
original_clinical_data <- global |> remove_labels()
```

# Data processing
The data set consists of thousands of features referring over 700 patients. The ones taken into account are sociodemographic, clinical and nutritional characteristics:

* **Sociodemographic data**
** `sexo`: gender ($0$ male, $1$ female).
** `edad`: age.
** `fumador_0m`: whether the patient is/was a smoker ($0$ no, $1$ yes).
** `trabajo_0m`: whether the patient is currently working ($0$ no, $1$ yes).
** `tipoest_0m`: educational level
** `numhog_0m`: number of people living in the same household.
** `numhij_0m`: number of children.
** `ecivil_0m`: marital status ($0$ single, $1$ married, $2$ divorced, $3$ widowed).
** `desenca_fa_0m`: whether the AF diagnosis was made after a cardiac event ($0$ no, $1$ yes).

* **Clinical data**
** `tipofa`: type of Atrial Fibrillation ($0$ paroxysmal,$1$ recurrent).
** `fec_fa_0m`: date of the first AF diagnosis.
** `fec_ablac_0m`: date of the first catheter ablation.
** `interv`: assigned cohort ($0$ control, $1$ intervention).
** `event18m_conkardia`: whether there has been AF recurrence after the blanking period and before $18$ months since the catheter ablation ($0$ no, $1$ yes).
** `ablacion_previa`: whether the patient has undergone catheter ablation previously to this trial ($0$ no, $1$ yes).
** `tiempo_fa_ablac`: years since the first AF diagnosis.
** `dilat_tot4_imp`: left atrial enlargement ($0$ normal, $1$ mild, $2$ moderate, $3$ severe).
** `bmi_autoref_0m`: Body Mass Index ($0$ normal weight, $25$ overweight, $30$ obese).
** `mettotal_0m`: MET (physical activity measurement).
** `hta_0m`: hypertension ($0$ no, $1$ yes).
** `diabetes2_0m`: type 2 diabetes ($0$ no, $1$ yes).
** `hipolipemiantes_0m`: whether the patient takes hypolipidemic medication ($0$ no, $1$ yes).
** `hipercol_0m`: whether the patient suffers from dyslipidemia ($0$ no, $1$ yes).
** `ictus_0m`: whether the patient has suffered from a stroke ($0$ no, $1$ yes).
** `ic_v00`: whether the patient has suffered from heart failure ($0$ no, $1$ yes).

* **Nutritional data**
** `p14_tot_0m:`: affinity to the Mediterranean diet ($0 - 14$).
** `energiat_0m`: total daily energy intake ($kcal$).
** `ac_oliva_0m`: olive oil daily intake ($g$).
** `ac_olivavir_0m`: extra virgin olive oil daily intake ($g$).
** `ac_orujo_0m`: pomace oil daily intake ($g$).
** `ac_maiz_0m`: corn oil daily intake ($g$).
** `ac_girasol_0m`: sunflower oil daily intake ($g$).
** `ac_soja_0m`: soy oil daily intake ($g$).
** `v_blanco_0m`: white wine daily intake ($g$).
** `v_tintojov_0m`: young red wine daily intake ($g$).
** `v_tintoaniej_0m`: aged red wine daily intake ($g$).
** `v_rosado_0m`: rose wine daily intake ($g$).
** `cervezas_0m`: beer daily intake ($g$).
** `alcoholg_0m`: total alcohol daily intake ($g$).

```{r}
clinical_data_selected <- original_clinical_data |> 
    select(
        # SOCIODEMOGRAPHIC DATA
        sexo,
        edad,
        fum_0m,
        trabajo_0m,
        tipoest_0m,
        numhog_0m,
        numhij_0m,
        ecivil_0m,
        
        # CLINICAL DATA
        code,
        interv,
        nodo,
        bmi_autoref_0m,
        mettotal_0m,
        diabetest2_0m,
        hipolipemiantes_0m,
        hipercol_0m,
        hta_0m,
        tipofa,
        ictus_0m,
        ic_v00,
        tiempo_fa_ablac,
        ablacion_previa,
        dilat_tot4,
        desenca_fa_0m,
        event18m_conkardia, # PREDICTED CLASS
        
        # NUTRITIONAL DATA (g/day)
        p14_tot_0m,
        energiat_0m,
        ac_oliva_0m,
        ac_olivavir_0m,
        ac_orujo_0m,
        ac_maiz_0m,
        ac_girasol_0m,
        ac_soja_0m,
        v_blanco_0m,
        v_tintojov_0m,
        v_tintoaniej_0m,
        v_rosado_0m,
        cervezas_0m,
        alcoholg_0m,
    )

head(clinical_data_selected)
```

Lets label the categorical data using factors:

```{r}
clinical_data_modified <- clinical_data_selected |> 
    mutate(
        sexo = factor(
            sexo, 
            levels = c(0, 1), 
            labels = c("male", "female"),
            ordered = FALSE
            ),
        
        fum_0m = factor(
            fum_0m,
            levels = c(0, 2, 1),
            labels = c("never", "current", "former"),
            ordered = TRUE
            ),
        
        trabajo_0m = factor(
            trabajo_0m,
            levels = seq(1, 5),
            labels = c(
                "full-time", "part-time", "houskeeper", "unemployed", "retired")
            ),
        
        tipoest_0m = factor(
            tipoest_0m,
            levels = c(0, 4, 3, 2, 1),
            labels = c(
                "none", "diploma", "college", "master", "doctorate"
            ),
            ordered = TRUE
        ),
        
        ecivil_0m = factor(
            ecivil_0m,
            levels = c(0, 1, 2, 3, 4),
            labels = c("single", "married", "widowed", "divorced", "other")
        ),
        
        tipofa = factor(
            tipofa,
            levels = c(0, 1),
            labels = c("paroxysmal", "persistent")
        ),
        
        interv = factor(
            interv,
            levels = c(0, 1),
            labels = c("control", "intervention")
        ),
        nodo = factor(
            nodo,
            levels = c(1, 2, 3, 4),
            labels = c("Pamplona", "Madrid", "Granada", "Alicante")
        ),
        
        event18m_conkardia = factor(
            event18m_conkardia,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        
        ablacion_previa = factor(
            ablacion_previa,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        
        hta_0m = factor(
            hta_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        
        dilat_tot4 = factor(
            dilat_tot4,
            levels = c(0, 1, 2, 3),
            labels = c("normal", "mild", "moderate", "severe"),
            ordered = TRUE
        ),
        
        diabetest2_0m = factor(
            diabetest2_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipolipemiantes_0m = factor(
            hipolipemiantes_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipercol_0m = factor(
            hipercol_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        ictus_0m = factor(
            ictus_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        ic_v00 = factor(
            ic_v00,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        desenca_fa_0m = factor(
            desenca_fa_0m,
            levels = c(1, 2, 3, 4, 6, 7, 8),
            labels = c(
                "physical exercise", "drinking", "emotion", "stress", 
                "lack of sleep", "heavy meals", "other"
                ),
            ordered = FALSE
        )
    )

head(clinical_data_modified)
```

```{r}
names(clinical_data_modified)
```

Let's modify and translate the variables' names to make them more understandable:

```{r}
new_names <- c(
    "gender", "age", "smoking_status", "working_status", "education_level", 
    "household_size", "children", "marital_status", "code", "cohort", "node", 
    "bmi", "total_met", "t2_diabetes", "hypolipidemics", "hypercholesterol", 
    "hbp", "fa_type", "stroke", "heart_failure", "fa_ablat_time", 
    "prev_ablation", "left_atrial_enlarg", "af_trigger", "af_recurrence", 
    "p14_score", "energy", "olive_oil", "ev_olive_oil", "pomace_oil", "corn_oil", 
    "sunflower_oil", "soy_oil", "white_wine", "young_red_wine", "aged_red_wine", 
    "rose_wine", "beer", "alcohol"
    )

clinical_data_processed <- clinical_data_modified |> setnames(
    old = names(clinical_data_modified), 
    new = new_names
    )
```

This is the primary data we will use to build the model. Lets save it:

```{r}
# Save it into the project's data directory
write_parquet(
    clinical_data_processed, 
    "data/intermediate/clinical_data.parquet"
    )

head(clinical_data_processed)
```
