---
title: "2.0- Exploratory Data Analysis"
output: html_document
---

# Configuration

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(skimr)
library(scales)
library(car)
```

# Load the data

```{r}
clinical_data <- read_parquet("data/intermediate/clinical_data.parquet")
head(clinical_data)
```

The `code` feature is merely descriptive and will not add information to the model, so we will drop it:

```{r}
clinical_data <- clinical_data |> select(-code)
```

# Global summary
Let's check the column's data types and whether there are many missing values:

```{r}
skim(clinical_data) |> summary()
```

There are $720$ rows and $39$ columns: $17$ categorical (the target class can be found among these) and $21$ numerical features. Let's check each type separately.

## Numerical features
Let's sum up the data:

```{r}
skim(clinical_data) |> yank("numeric") |> 
  select(-c(starts_with("p"), hist, complete_rate))
```

The only feature with missing values is `af_trigger`. Given than they represent the majority of the records are missing values, we should either find out whether this is because the trigger is unknown or the data is truly missing. In the first case, we can impute the missing values with a new category, "unknown". However, if the data is truly missing, the feature should be dropped from the data set.

## Categorical features

Let's sum up the data:

```{r}
skim(clinical_data) |> yank("factor") |> select(-c(complete_rate, ordered))
```

There are $17$ missing values in the `hypolipidemics` feature and $37$ in the `left_atrial_enlarg`. We will decide on how to deal with these missing values later on.

# Statistical analysis
Firstly, let's check how the data is distributed.

## Numerical features

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free", ncol = 4) +
    theme_minimal()
```

Most of the numerical features don't show a normal distribution. Apart from this, the `corn_oil`, `pomace_oil`, `rose_wine` and `soy_oil` distributions show that all their values are zero. Let's check these features individually:

```{r}
clinical_data |> select(corn_oil) |> table()
```

```{r}
clinical_data |> select(pomace_oil) |> table()
```

```{r}
clinical_data |> select(rose_wine) |> table()
```

```{r}
clinical_data |> select(soy_oil) |> table()
```

Since the `corn_oil` and the `pomace_oil` features are empty, we will drop them from the data set:

```{r}
clinical_data <- clinical_data |> 
  select(-c(corn_oil, pomace_oil))
```

Let's generate the new histogram with the remaining features:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value)) +
    geom_histogram(bins = 10, color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal()
```

Now, let's check how does the distribution of each feature behaves when we stratify the data by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric), af_recurrence) |> 
    pivot_longer(
        cols = -af_recurrence,
        names_to = "feature",
        values_to = "value"
    ) |> 
    ggplot(aes(x = af_recurrence, y = value)) +
    geom_boxplot() +
    scale_y_continuous(trans = pseudo_log_trans(base = 10)) +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal()
```

## Categorical features

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.factor)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value",
        values_transform = list(value = as.character)
        ) |> 
    ggplot(aes(x = value)) +
    geom_bar(color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free") + 
    labs(x = "", y = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Most of these categories are unbalanced, including the target class. Now let's see how this changes when we stratify the data by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.factor)) |> 
    pivot_longer(
        cols = -af_recurrence,
        names_to = "feature",
        values_to = "value",
        values_transform = list(value = as.character)
        ) |> 
    ggplot(aes(x = value, fill = af_recurrence)) +
    geom_bar(color = "black", position = "fill") +
    facet_wrap(~ feature, scales = "free") + 
    labs(x = "", y = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Multicolinearity
Finally, let's check the correlation between features to drop those that are highly correlated. To address numerical features' colinearity, we will use the **VIF**.

```{r}
# Convert the target class to numeric for the VIF calculation
temp_clinical_data <- clinical_data |> 
  mutate(af_recurrence = as.numeric(af_recurrence))

# Fit a linear model with all the features
model <- lm(
    af_recurrence ~ ., 
    data = temp_clinical_data
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The categorical features aren't correlated at all. On the other hand, the alcoholic drinks are somehow correlated with each other, specially with the `alcohol` feature. Therefore we will drop the `alcohol` feature from the data set and calculate again the VIFs:

```{r}
# drop the alcohol feature
temp_df <- temp_clinical_data |> 
    select(-alcohol)

# fit a linear model with the remaining numeric features
model <- lm(
    af_recurrence ~ ., 
    data = temp_df
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What happens if we keep the `alcohol` feature and drop the ones referring to specific alcoholic drinks?

```{r}
# drop the alcoholic beverages features
temp_df <- temp_clinical_data |> 
    select(-c(white_wine, young_red_wine, aged_red_wine, beer))

# fit a linear model with the remaining numeric features
model <- lm(
    af_recurrence ~ ., 
    data = temp_df
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since the maximum VIF is slightly lower than in the first scenario, we keep the `alcohol` feature and drop the ones referring to specific alcoholic drinks. Apart from this, the data set's dimensionality is reduced.

```{r}
clinical_data <- clinical_data |> 
    select(-c(white_wine, young_red_wine, aged_red_wine, rose_wine, beer))
```

# Save progress
Let's save this data set for the next steps:

```{r}
# Save it into the project's data directory
write_parquet(
    clinical_data, 
    "data/intermediate/clinical_data_explored.parquet"
    )

head(clinical_data)
```
