---
title: "2.0- Exploratory Data Analysis"
output: html_document
---

# Packages

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(skimr)
library(scales)
library(car)
```

# Load the data

```{r}
# full data
clinical_data <- read_parquet("data/intermediate/clinical_data_clean.parquet")

head(clinical_data)
```

The `id` feature is merely descriptive and will not add information to the model, so we will drop it:

```{r}
clinical_data <- clinical_data |> select(-id)
```

# Global summary
Let's check the column's data types and whether there are many missing values:

```{r}
skim(clinical_data) |> summary()
```

There are 720 rows and 30 columns: 11 categorical (one of them is the predicted class) and 19 numerical variables. Let's check each type separately. 

## Numerical features
Let's sum up the data:

```{r}
skim(clinical_data) |> yank("numeric") |> 
  select(-c(starts_with("p"), hist, complete_rate))
```

There are no missing values in the numerical features.

## Categorical features

Let's sum up the data:

```{r}
skim(clinical_data) |> yank("factor") |> select(-c(complete_rate, ordered))
```

There are 17 missing values in the `hypolipidemic` feature.

```{r}
clinical_data |> select(hypolipidemic) |> table()
```

Since the class is unbalanced, we will decide later on how to deal with these missing values.

# Statistical analysis
Firstly, let's check how the data is distributed.

## Numerical features

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value)) +
    geom_histogram(bins = 10, color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal()
```

Most of the numerical features don't show a normal distribution. Apart from this, the `corn_oil`, `pomace_oil`, `rose_wine` and `soy_oil` distributions show that all their values are zero. Let's check these features individually:

```{r}
clinical_data |> select(corn_oil) |> table()
```

```{r}
clinical_data |> select(pomace_oil) |> table()
```

```{r}
clinical_data |> select(rose_wine) |> table()
```

```{r}
clinical_data |> select(soy_oil) |> table()
```

Since all these features are empty or almost empty, we will drop them from the data set:

```{r}
clinical_data <- clinical_data |> 
  select(-c(corn_oil, pomace_oil, rose_wine, soy_oil))
```

Let's generate the new histogram with the remaining features:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value)) +
    geom_histogram(bins = 10, color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal()
```

Now, let's check how does the distribution of each feature behaves when we stratify the data by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.numeric), cardiac_event) |> 
    pivot_longer(
        cols = -cardiac_event,
        names_to = "feature",
        values_to = "value"
    ) |> 
    ggplot(aes(x = cardiac_event, y = value)) +
    geom_boxplot() +
    scale_y_continuous(trans = pseudo_log_trans(base = 10)) +
    facet_wrap(~ feature, scales = "free") +
    theme_minimal()
```

## Categorical features

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.factor)) |> 
    pivot_longer(
        cols = everything(),
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value)) +
    geom_bar(color = "black", fill = "cyan") +
    facet_wrap(~ feature, scales = "free") + 
    labs(x = "", y = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Most of these categories are unbalanced, including the target class. Now let's see how this changes when we stratify the data by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
    select(where(is.factor)) |> 
    pivot_longer(
        cols = -cardiac_event,
        names_to = "feature",
        values_to = "value"
        ) |> 
    ggplot(aes(x = value, fill = cardiac_event)) +
    geom_bar(color = "black", position = "fill") +
    facet_wrap(~ feature, scales = "free") + 
    labs(x = "", y = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since most of the missing values in the `hypolipidemic` feature belong to the majority category, we will delete those records:

```{r}
clinical_data <- clinical_data |> drop_na()
```

# Multicolinearity
Finally, let's check the correlation between features to drop those that are highly correlated. To address numerical features' colinearity, we will use the **VIF**.

```{r}
# Fit a linear model with all the features
model <- glm(
    cardiac_event ~ ., 
    data = clinical_data, 
    family = binomial
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The categorical features aren't correlated at all. On the other hand, the alcoholic drinks are somehow correlated with each other, specially with the `alcohol` feature. Therefore we will drop the `alcohol` feature from the data set and calculate again the VIFs:

```{r}
# drop the alcohol feature
temp_df <- clinical_data |> 
    select(-alcohol)

# fit a linear model with the remaining numeric features
model <- glm(
    cardiac_event ~ ., 
    data = temp_df, 
    family = binomial
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What happens if we keep the `alcohol` feature and drop the ones referring to specific alcoholic drinks?

```{r}
# drop the alcoholic beverages features
temp_df <- clinical_data |> 
    select(
        -c(white_wine, young_red_wine, aged_red_wine, beer)
        )

# fit a linear model with the remaining numeric features
model <- glm(
    cardiac_event ~ ., 
    data = temp_df, 
    family = binomial
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since the maximum VIF is the same, we keep the `alcohol` feature and drop the ones referring to specific alcoholic drinks. This way, we reduce the data set's dimensionality as we eliminate multicollinearity.

```{r}
clinical_data <- clinical_data |> 
    select(
        -c(white_wine, young_red_wine, aged_red_wine, beer)
        )
```

# Final data set
Finally, let's save the data set with the selected features:

```{r}
clinical_data |> 
  write_parquet("data/processed/clinical_data_selected_features.parquet")

head(clinical_data)
```
