---
title: "2.0- Exploratory Data Analysis"
output: html_document
---

# Configuration

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(skimr)
library(scales)
library(car)

source("src/data_wrangling/data_analysis.R")
```

# Load the data

```{r}
clinical_data <- read_parquet("data/intermediate/clinical_data.parquet")
head(clinical_data)
```

# Global summary
Let's check the column's data types and whether there are many missing values:

```{r}
skim(clinical_data) |> summary()
```

There are $720$ rows and $34$ columns: $20$ categorical (the target class can be found among these) and $14$ numerical features. Let's check each type separately.

## Numerical features
Let's sum up the data:

```{r}
clinical_data |> select(-code) |> skim() |> yank("numeric") |> 
  select(-c(starts_with("p"), hist, complete_rate))
```

Since half of the data in the `LAD`, `LAA` and `LAV` columns is missing, we will drop those features:

```{r}
clinical_data <- clinical_data |>
  select(-c(LAD, LAA, LAV))
```

The remaining columns with missing values are: `waist_to_height`, `glucose`, `HDL`, and `triglicerides`. Let's check if these missing values are concentrated in a specific class of the target variable:

```{r}
clinical_data |> 
  filter(is.na(waist_to_height)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

```{r}
clinical_data |> 
  filter(is.na(glucose)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

```{r}
clinical_data |> 
  filter(is.na(HDL)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

```{r}
clinical_data |> 
  filter(is.na(triglicerides)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

```{r}
clinical_data |> 
  filter(is.na(waist_to_height) & is.na(glucose) & is.na(HDL) & is.na(triglicerides)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

## Categorical features

Let's sum up the data:

```{r}
skim(clinical_data) |> yank("factor") |> select(-c(complete_rate, ordered))
```

There are $17$ missing values in the `hypolipidemic_meds` feature. Let's check if these missing values are concentrated in a specific class of the target variable:

```{r}
clinical_data |> 
  filter(is.na(hypolipidemic_meds)) |>
  group_by(AF_recurrence) |>
  summarise(count = n())
```

# Statistical analysis
Firstly, let's check how is the data distributed.

## Numerical features

```{r, fig.width=10, fig.height=8}
clinical_data |> 
  select(-code) |> 
  plot_global_numeric_distribution()
```

Now, let's check the distribution stratified by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> 
  select(-code) |> 
  plot_stratified_numeric_distribution(target_var = "AF_recurrence")

```

## Categorical features

```{r, fig.width=16, fig.height=12}
clinical_data |> select(-code) |> 
  plot_global_categorical_distribution()
```

Most of these categories are imbalanced, including the target class. Now let's see how this changes when we stratify the data by the target class:

```{r, fig.width=10, fig.height=16}
clinical_data |> select(-code) |> 
  plot_stratified_categorical_distribution(target_var = "AF_recurrence")
```

## Multicolinearity
Let's check the correlation between features to drop those that are highly correlated. 

### Numeric features
Firstly, let's take a look at the correlation matrix:

```{r}
clinical_data |> select(-code) |> plot_corr_matrix(threshold = 0.5)
```

Since the waist to height ratio is a better predictor of obesity, a risk factor for AF recurrence, than the waist circumference, we will drop the `BMI` feature:

```{r}
clinical_data <- clinical_data |> select(-BMI)
```


### Categorical features
To address categorical features' colinearity, we will use **Cramer's V**.

```{r}
clinical_data |> 
  plot_corr_matrix(dtype = "cat", threshold = 0.5) |> 
  suppressWarnings()
```

There is a perfect correlation between the `heart_failure` and `LVEF` features. Since the latter is more informative, we will drop the `heart_failure` feature:

```{r}
clinical_data <- clinical_data |> select(-heart_failure)
```

### Global correlation
To address global colinearity, we will use the **VIF**.

```{r}
# Convert the target class to numeric for the VIF calculation
temp_clinical_data <- clinical_data |> select(-code) |> 
  mutate(AF_recurrence = as.numeric(AF_recurrence))

# Fit a linear model with all the features
model <- lm(
    AF_recurrence ~ ., 
    data = temp_clinical_data
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since $VIF < 5$ for all of the features, there is no significant multicollinearity among anyone of them.

# Missing values

# Save progress

```{r}
# Save the data
write_parquet(
    clinical_data, 
    "data/clean/clinical_data.parquet"
    )

head(clinical_data)
```
