---
title: "Data cleaning"
output: html_document
---

# Packages

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(labelled)
library(arrow)
library(data.table)

source("src/data_wrangling/data_cleaning.R")
```

# Load the data
This data set stores clinical information from patients in the PREDIMAR study. The aim of this project is to build a machine learning model capable of predicting AF recurrence based on medical data.

Let's load the data set:

```{r echo = FALSE}
global <- read_dta("data/raw/predimar_miguelgomez.dta")
```

Since it is a STATA file, the labels need to be removed:

```{r}
original_clinical_data <- global |> remove_labels()
```

# Features selection
The data set consists of thousands of features referring over 700 patients. The ones taken into account are sociodemographic, clinical and nutritional characteristics:

* **Sociodemographic data**
** `sexo`: gender ($0$ male, $1$ female).
** `edad`: age.
** `fum_0m`: whether the patient is/was a smoker ($0$ no, $1$ yes).

* **Clinical data**
** `code`: patient tracking identifier.
** `nodo`: assigned hospital center.
** `interv`: assigned cohort ($0$ control, $1$ intervention).
** `bmi_autoref_0m`: Body Mass Index ($0$ normal weight, $25$ overweight, $30$ obese).
** `cintura_0m`: waist circumference ($cm$).
** `altura_0m:`height ($cm$).
** `mettotal_0m`: Met (physical activity measurement).
** `glu_new0m`: glucose level ($mg/dL$).
** `diabetest1_0m`:type 1 diabetes ($0$ no, $1$ yes).
** `diabetest2_0m`:type 2 diabetes ($0$ no, $1$ yes).
** `hdl_new0m`: HDL cholesterol level ($mg/dL$).
** `trig_new0m`: triglycerides level ($mg/dL$).
** `hipercol_0m`: whether the patient suffers from dyslipidemia ($0$ no, $1$ yes).
** `hipolipemiantes_0m`: whether the patient takes hypolipidemic medication ($0$ no, $1$ yes).
** `saos_0m`: whether the patient suffers from sleep apnea ($0$ no, $1$ yes). 
** `insurenal_0m`: whether the patient suffers from renal insufficiency ($0$ no, $1$ yes).
** `hta_0m`: hypertension ($0$ no, $1$ yes).
** `epoc_0m`: whether the patient suffers from chronic obstructive pulmonary disease ($0$ no, $1$ yes).
** `ictus_0m`: whether the patient has suffered from a stroke ($0$ no, $1$ yes).
** `ic_v00`: whether the patient has suffered from heart failure ($0$ no, $1$ yes).
** `miocardiopat_0m_imp`: whether the patient suffers from cardiomyopathy ($0$ no, $1$ yes).
** `faa_0m_new`: whether the patient is having antiarrhythmic drugs ($0$ no, $1$ yes).
** `dilat_tot4_imp`: left atrial enlargement ($0$ normal, $1$ mild, $2$ moderate, $3$ severe).
** `eco_diamai_0m`: left atrial diameter ($cm$).
** `eco_areaai_0m`: left atrial area ($cm^2)$.
** `eco_volumenai_0m`: left atrial volume ($mL)$.
** `eco_fe_0m`: left atrial ejection fraction ($\%$).
** `tipofa`: type of Atrial Fibrillation ($0$ paroxysmal, $1$ persistent).
** `tiempo_fa_ablac`: years since the first AF diagnosis.
** `ablacion_previa`: whether the patient has undergone catheter ablation previously to this trial ($0$ no, $1$ yes).
** `event_blank`: whether there has been AF recurrence during the blanking period ($0$ no, $1$ yes).
** `event18m_conkardia`: whether there has been AF recurrence between $3$ and $18$ months since the catheter ablation ($0$ no, $1$ yes).

* **Nutritional data**
** `p14_tot_0m`: nutritional score measuring affinity to the mediterranean diet ($0 - 14$).
** `alcoholg_0m`: total alcohol daily intake ($g$).

```{r}
clinical_data_selected <- original_clinical_data |> 
    select(
        #=======================================================================
        # SOCIODEMOGRAPHIC DATA
        #=======================================================================
        sexo,
        edad,
        fum_0m,
        
        #=======================================================================
        # CLINICAL DATA
        #=======================================================================
        # Tracking & Trial Context
        code,
        nodo,
        interv,
        
        # Lifestyle & General Comorbidities
        bmi_autoref_0m,
        cintura_0m,
        altura_0m,
        mettotal_0m,
        glu_new0m,
        diabetest1_0m,
        diabetest2_0m,
        hdl_new0m,
        trig_new0m,
        hipercol_0m,
        hipolipemiantes_0m,
        saos_0m,
        insurenal_0m,
        hta_0m,
        epoc_0m,
        ictus_0m,
        
        # Specific Cardiac Conditions & Echo Structural Measurements
        ic_v00,
        miocardiopat_0m_imp,
        faa_0m_new,
        dilat_tot4_imp,
        eco_diamai_0m,
        eco_areaai_0m,
        eco_volumenai_0m,
        eco_fe_0m,
        
        # Atrial Fibrillation History & Dynamics
        tipofa,
        tiempo_fa_ablac,
        ablacion_previa,
        event_blank,
        
        # Target class
        event18m_conkardia
    )

head(clinical_data_selected)
```

# Data cleaning
Lets label the categorical data using factors:

```{r}
clinical_data_modified <- clinical_data_selected |> 
    mutate(
        # ======================================================================
        # SOCIODEMOGRAPHIC DATA
        # ======================================================================
        sexo = factor(
            sexo, 
            levels = c(0, 1), 
            labels = c("male", "female"),
            ordered = TRUE
        ),
        
        fum_0m = factor(
            fum_0m,
            levels = c(0, 2, 1),
            labels = c("never", "current", "former"),
            ordered = TRUE
        ),
        
        # ======================================================================
        # CLINICAL DATA
        # ======================================================================
        # Tracking & Trial Context
        nodo = factor(
            nodo,
            levels = c(1, 2, 3, 4),
            labels = c("Pamplona", "Madrid", "Granada", "Alicante"),
            ordered = FALSE
        ),
        
        interv = factor(
            interv,
            levels = c(0, 1),
            labels = c("control", "intervention"),
            ordered  = TRUE
        ),

        # Lifestyle & General Comorbidities
        diabetest1_0m = factor(
            diabetest1_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        diabetest2_0m = factor(
            diabetest2_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipercol_0m = factor(
            hipercol_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipolipemiantes_0m = factor(
            hipolipemiantes_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        saos_0m = factor(
            saos_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        insurenal_0m = factor(
            insurenal_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hta_0m = factor(
            hta_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        epoc_0m = factor(
            epoc_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        ictus_0m = factor(
            ictus_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        # Specific Cardiac Conditions & Echo Measurements
        ic_v00 = factor(
            ic_v00,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        miocardiopat_0m_imp = factor(
            miocardiopat_0m_imp,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        faa_0m_new = factor(
            faa_0m_new,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        dilat_tot4_imp = factor(
            dilat_tot4_imp,
            levels = c(0, 1, 2, 3),
            labels = c("normal", "mild", "moderate", "severe"),
            ordered = TRUE
        ),

        eco_fe_0m = factor(
            eco_fe_0m,
            levels = c(0, 1, 2, 3),
            labels = c("normal", "slightly abnormal",
                       "moderately abnormal", "severely abnormal"),
            ordered = TRUE
        ),
        
        # =====================================================================
        # CLINICAL DATA: Atrial Fibrillation History & Dynamics
        # =====================================================================
        tipofa = factor(
            tipofa,
            levels = c(0, 1),
            labels = c("persistent", "paroxysmal")
        ),
        
        ablacion_previa = factor(
            ablacion_previa,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        event_blank = factor(
            event_blank,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        # TARGET CLASS
        event18m_conkardia = factor(
            event18m_conkardia,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        )
    ) |> 
    
    # Reorder the factor levels appropriately
    mutate(
        tipofa = fct_relevel(tipofa, "paroxysmal")
    )

head(clinical_data_modified)
```

Display the current names of the columns:

```{r}
names(clinical_data_modified)
```

Let's modify them:

```{r}
new_names <- c(
    "sex", "age", "smoking_status", "code", "center", "intervention", "BMI", 
    "waist_circumference", "height", "Met", "glucose", "type1_diabetes", 
    "type2_diabetes", "HDL", "triglicerides", "hypercholesterolemia", 
    "hypolipidemic_meds", "OSA", "renal_insuf", "hypertension", "COPD", 
    "stroke", "heart_failure", "cardiomyopathy", "antirrythmic_meds", 
    "LA_enlargment", "LAD", "LAA", "LAV", "LVEF", "AF_type", 
    "AF_ablation_time", "previous_ablation", "ERAF", 
    "AF_recurrence"
    )

clinical_data_processed <- clinical_data_modified |> setnames(
    old = names(clinical_data_modified), 
    new = new_names
    )
```

# Feature engineering
As an alternative to the `BMI` feature, we can compute the waist-to-height ratio to assess obesity:

```{r}
clinical_data_processed <- clinical_data_processed |> 
    mutate(
        waist_height_ratio = waist_circumference / height,
        .after = BMI
    ) |> 
    select(-c(waist_circumference, height))
```

We can compute the HATCH score for each patient:
```{r}
clinical_data_processed <- compute_hatch_score(clinical_data_processed)
```


# Save data
Save the data:

```{r}
# Full data set
write_parquet(
    clinical_data_processed, 
    "data/intermediate/clinical_data_full.parquet"
    )

# Subset for hatch score
clinical_data_processed |> 
    select(AF_recurrence, hatch_score, age, heart_failure, hypertension, COPD, stroke, OSA) |> 
    write_parquet("data/clean/clinical_data_hatch.parquet")

head(clinical_data_processed)
```
