---
title: "1.0-Data processing"
output: html_document
---

# Packages

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(labelled)
library(arrow)
library(data.table)
```

# Load the data
This data set stores clinical information from patients in the PREDIMAR study. The aim of this project is to build a machine learning model capable of predicting AF recurrence based on medical data.

Let's load the data set:

```{r echo = FALSE}
global <- read_dta("data/raw/predimar_miguelgomez.dta")
```

Since it is a STATA file, the labels need to be removed:

```{r}
original_clinical_data <- global |> remove_labels()
```

# Data processing
The data set consists of thousands of features referring over 700 patients. The ones taken into account are sociodemographic, clinical and nutritional characteristics:

* **Sociodemographic data**
** `sexo`: gender ($0$ male, $1$ female).
** `edad`: age.
** `fum_0m`: whether the patient is/was a smoker ($0$ no, $1$ yes).
** `trabajo_0m`: whether the patient is currently working ($0$ no, $1$ yes).
** `tipoest_0m`: educational level.
** `ecivil_0m`: marital status ($0$ single, $1$ married, $2$ divorced, $3$ widowed).
** `numhog_0m`: number of people living in the same household.
** `numhij_0m`: number of children.

* **Clinical data**
** `code`: patient tracking identifier.
** `nodo`: assigned hospital node/center.
** `interv`: assigned cohort ($0$ control, $1$ intervention).
** `bmi_autoref_0m`: Body Mass Index ($0$ normal weight, $25$ overweight, $30$ obese).
** `mettotal_0m`: MET (physical activity measurement).
** `hta_0m`: hypertension ($0$ no, $1$ yes).
** `diabetest2_0m`: type 2 diabetes ($0$ no, $1$ yes).
** `hipercol_0m`: whether the patient suffers from dyslipidemia ($0$ no, $1$ yes).
** `hipolipemiantes_0m`: whether the patient takes hypolipidemic medication ($0$ no, $1$ yes).
** `saos_0m`: whether the patient suffers from sleep apnea ($0$ no, $1$ yes). 
** `ictus_0m`: whether the patient has suffered from a stroke ($0$ no, $1$ yes).
** `ic_v00`: whether the patient has suffered from heart failure ($0$ no, $1$ yes).
** `miocardiopat_0m_imp`: whether the patient suffers from cardiomyopathy ($0$ no, $1$ yes).
** `dilat_tot4`: left atrial enlargement ($0$ normal, $1$ mild, $2$ moderate, $3$ severe).
** `eco_diamai_0m`: left atrial diameter ($cm$).
** `eco_areaai_0m`: left atrial area in ($cm^2)$.
** `eco_volumenai_0m`: left atrial volume ($mL)$.
** `tipofa`: type of Atrial Fibrillation ($0$ paroxysmal, $1$ recurrent).
** `desenca_fa_0m`: whether the AF diagnosis was made after a cardiac event ($0$ no, $1$ yes).
** `tiempo_fa_ablac`: years since the first AF diagnosis.
** `ablacion_previa`: whether the patient has undergone catheter ablation previously to this trial ($0$ no, $1$ yes).
** `event_blank`: whether there has been AF recurrence during the blanking period ($0$ no, $1$ yes).
** `event18m_conkardia`: whether there has been AF recurrence after the blanking period and before $18$ months since the catheter ablation ($0$ no, $1$ yes).

* **Nutritional data**
** `p14_tot_0m`: affinity to the Mediterranean diet ($0 - 14$).
** `alcoholg_0m`: total alcohol daily intake ($g$).

```{r}
clinical_data_selected <- original_clinical_data |> 
    select(
        # SOCIODEMOGRAPHIC DATA
        sexo,
        edad,
        fum_0m,
        trabajo_0m,
        tipoest_0m,
        ecivil_0m,
        numhog_0m,
        numhij_0m,
        
        # CLINICAL DATA: Tracking & Trial Context
        code,
        nodo,
        interv,
        
        # CLINICAL DATA: Lifestyle & General Comorbidities
        bmi_autoref_0m,
        mettotal_0m,
        hta_0m,
        diabetest2_0m,
        hipercol_0m,
        hipolipemiantes_0m,
        saos_0m,
        ictus_0m,
        
        # CLINICAL DATA: Specific Cardiac Conditions & Echo Structural Measurements
        ic_v00,
        miocardiopat_0m_imp,
        dilat_tot4,
        eco_diamai_0m,
        eco_areaai_0m,
        eco_volumenai_0m,
        
        # CLINICAL DATA: Atrial Fibrillation History & Dynamics
        tipofa,
        desenca_fa_0m,
        tiempo_fa_ablac,
        ablacion_previa,
        event_blank,
        
        # PREDICTED CLASS / OUTCOME
        event18m_conkardia,
        
        # NUTRITIONAL DATA
        p14_tot_0m,
        alcoholg_0m
    )

head(clinical_data_selected)
```

Lets label the categorical data using factors:

```{r}
clinical_data_modified <- clinical_data_selected |> 
    mutate(
        # =====================================================================
        # SOCIODEMOGRAPHIC DATA
        # =====================================================================
        sexo = factor(
            sexo, 
            levels = c(0, 1), 
            labels = c("male", "female"),
            ordered = FALSE
        ),
        
        fum_0m = factor(
            fum_0m,
            levels = c(0, 2, 1),
            labels = c("never", "current", "former"),
            ordered = TRUE
        ),
        
        trabajo_0m = factor(
            trabajo_0m,
            levels = seq(1, 5),
            labels = c("full-time", "part-time", "houskeeper", "unemployed", "retired")
        ),
        
        tipoest_0m = factor(
            tipoest_0m,
            levels = c(0, 4, 3, 2, 1),
            labels = c("none", "diploma", "college", "master", "doctorate"),
            ordered = TRUE
        ),
        
        ecivil_0m = factor(
            ecivil_0m,
            levels = c(0, 1, 2, 3, 4),
            labels = c("single", "married", "widowed", "divorced", "other")
        ),
        
        # =====================================================================
        # CLINICAL DATA: Tracking & Trial Context
        # =====================================================================
        nodo = factor(
            nodo,
            levels = c(1, 2, 3, 4),
            labels = c("Pamplona", "Madrid", "Granada", "Alicante")
        ),
        
        interv = factor(
            interv,
            levels = c(0, 1),
            labels = c("control", "intervention")
        ),
        
        # =====================================================================
        # CLINICAL DATA: Lifestyle & General Comorbidities
        # =====================================================================
        hta_0m = factor(
            hta_0m,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        
        diabetest2_0m = factor(
            diabetest2_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipercol_0m = factor(
            hipercol_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        hipolipemiantes_0m = factor(
            hipolipemiantes_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        saos_0m = factor(
            saos_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        ictus_0m = factor(
            ictus_0m,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        # =====================================================================
        # CLINICAL DATA: Specific Cardiac Conditions & Echo Measurements
        # =====================================================================
        ic_v00 = factor(
            ic_v00,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        miocardiopat_0m_imp = factor(
            miocardiopat_0m_imp,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        dilat_tot4 = factor(
            dilat_tot4,
            levels = c(0, 1, 2, 3),
            labels = c("normal", "mild", "moderate", "severe"),
            ordered = TRUE
        ),
        
        # =====================================================================
        # CLINICAL DATA: Atrial Fibrillation History & Dynamics
        # =====================================================================
        tipofa = factor(
            tipofa,
            levels = c(0, 1),
            labels = c("paroxysmal", "persistent")
        ),
        
        desenca_fa_0m = factor(
            desenca_fa_0m,
            levels = c(1, 2, 3, 4, 6, 7, 8),
            labels = c("physical exercise", "drinking", "emotion", "stress", 
                       "lack of sleep", "heavy meals", "other"),
            ordered = FALSE
        ),
        
        ablacion_previa = factor(
            ablacion_previa,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        
        event_blank = factor(
            event_blank,
            levels = c(0, 1),
            labels = c("no", "yes"),
            ordered = TRUE
        ),
        
        # =====================================================================
        # PREDICTED CLASS / OUTCOME
        # =====================================================================
        event18m_conkardia = factor(
            event18m_conkardia,
            levels = c(0, 1),
            labels = c("no", "yes")
        )
    )

head(clinical_data_modified)
```

```{r}
names(clinical_data_modified)
```

Let's modify and translate the variables' names to make them more understandable:

```{r}
new_names <- c(
    "gender", "age", "smoking_status", "working_status", "education_level", 
    "marital_status", "household_size", "children", "code", "center", 
    "intervention", "BMI", "MET", "hypertension", "type2_diabetes", 
    "hypercholesterolemia", "hypolipidemic_medication", "OSA", "stroke", 
    "heart_failure", "cardiomyopathy", "atrial_enlargement", "LAD", "LAA", 
    "LAV", "AF_type", "AF_trigger", "AF_ablation_time", "previous_ablation", 
    "AF_recurrence_blank_period", "AF_recurrence", "p14_score", "alcohol_intake"
    )

clinical_data_processed <- clinical_data_modified |> setnames(
    old = names(clinical_data_modified), 
    new = new_names
    )
```

This is the primary data we will use to build the model. Lets save it:

```{r}
# Save it into the project's data directory
write_parquet(
    clinical_data_processed, 
    "data/intermediate/clinical_data.parquet"
    )

head(clinical_data_processed)
```
