---
title: "Clinical Data Analysis"
output: html_document
---

# Configuration

```{r}
library(here)
library(tidyverse)
library(arrow)
library(skimr)
library(scales)
library(car)

source(here("src", "data_wrangling", "data_analysis.R"))
source(here("src", "data_wrangling", "plots.R"))

data_dir <- here("data", "intermediate", "clinical_data.parquet")
output_dir <- here("results", "eda")
```

# Load the data

```{r}
clinical_data <- read_parquet(data_dir)
head(clinical_data)
```

# Global summary
Let's check the column's data types and whether there are many missing values:

```{r}
skim(clinical_data) |> summary()
```

There are $720$ rows and $36$ columns: $22$ categorical (the target class can be found among these) and $14$ numeric features. Let's check each type separately.

## Numeric features

Let's sum up the data:

```{r}
# Numeric features summary
num_features_summary <- clinical_data |> 
  select(-code) |> 
  
  # Compute summary statistics for numeric features
  skim() |> yank("numeric") |> 
  
  select(c(skim_variable, mean, sd, complete_rate)) |> 
  rename(feature = skim_variable) |> 
  
  # Replace the rate of complete values with the rate of missing values
  mutate(
    missing_rate = 1 - complete_rate,
    complete_rate = NULL
    )

print(num_features_summary, digits = 4)

# Save results
output_file <- here(output_dir, "num_features_summary.csv")

num_features_summary |> 
  write.csv(output_file, row.names = FALSE)
```

We will drop the features with $missing \ rate > 25 \%$:

```{r}
num_features_summary |> 
  filter(missing_rate > 0.25) |> 
  select(c(feature, missing_rate))
```

Therefore, we will drop these features:

```{r}
clinical_data <- clinical_data |>
  select(-c(LAD, LAA, LAV))
```

## Categorical features
Let's sum up the data:

```{r}
# Categorical features summary
cat_features_summary <- clinical_data |> 
  select(-code) |> 
  
  # Compute summary statistics for categorical features
  skim() |> 
  yank("factor") |> 
  select(-c(complete_rate, ordered, top_counts)) |> 
  rename(feature = skim_variable) |> 
  mutate(
    missing_rate = n_missing/nrow(clinical_data),
    n_missing = NULL
    )

print(cat_features_summary, digits = 4, n = nrow(cat_features_summary))

# Save results
output_file <- here(output_dir, "cat_features_summary.csv")

cat_features_summary |> 
  write.csv(output_file, row.names = FALSE)
```

Using the same criteria as for the numeric features, we will drop those categorical features with $missing \ rate > 25 \%$:

```{r}
cat_features_summary |> 
  filter(missing_rate > 0.25) |> 
  select(c(feature, missing_rate))
```

Therefore, we will drop these features:

```{r}
clinical_data <- clinical_data |>
  select(-c(eGFR_MDRD, eGFR_CG, ischemic_cardiomyopathy))
```

# Missing values
Let's check the missing values by column:

```{r}
# Data frame
clinical_data_missingness <- clinical_data |> 
  skim() |> 
  select(c(skim_variable, complete_rate)) |> 
  rename(feature = skim_variable) |> 
  
  # Replace the rate of complete values with the rate of missing values
  mutate(
    missing_rate = 1 - complete_rate,
    complete_rate = NULL
    )

# Create the plot
NAs_by_col <- plot_column_missingness(clinical_data_missingness)

# Visualize the plot
NAs_by_col
```

We will delete those records and features where $missing\_rate > 25 \%$:

```{r}
# Create the plot
NAs_by_row <- plot_row_missingness(clinical_data)

# Visualize the plot
NAs_by_row
```

Since there are no records with $missing \ rate > 25 \%$), we will keep all of them.

# Statistical analysis

## Numeric features

```{r}
# Create the plot
global_numeric_distribution <- clinical_data |> 
  select(-c(code, hatch_score)) |> 
  plot_global_numeric_distribution()

# Show the plot
global_numeric_distribution
```

Numeric features distribution stratified by the target class:

```{r}
# Create the plot
numeric_distribution_by_AF_recurrence <- clinical_data |> 
  select(-c(code, hatch_score)) |> 
  plot_stratified_numeric_distribution(target_var = "AF_recurrence")

# Show score
numeric_distribution_by_AF_recurrence
```

Numeric features distribution stratified by the intervention feature:

```{r}
# Create the plot
numeric_distribution_by_intervention <- clinical_data |> 
  select(-c(code, hatch_score)) |> 
  plot_stratified_numeric_distribution(target_var = "intervention")

# Show the plot
numeric_distribution_by_intervention
```

## Categorical features

Categorical features distribution:

```{r}
# Create the plot
categorical_distribution <- clinical_data |> 
  plot_single_categorical_distribution(var_name = "LA_enlargment")

# Show the plot
categorical_distribution
```

Categorical features distribution stratified by the target class:

```{r}
# Create the plot
categorical_distribution_by_AF_recurrence <- clinical_data |> 
  plot_single_stratified_categorical(var_name = "LA_enlargment", target_var = "AF_recurrence")

# Show the plot
categorical_distribution_by_AF_recurrence
```

Categorical features distribution stratified by the intervention variable:

```{r}
# Create the plot
categorical_distribution_by_intervention <- clinical_data |> 
  plot_single_stratified_categorical(var_name = "AF_type", target_var = "intervention")

# Show the plot
categorical_distribution_by_intervention
```

## Multicolinearity
Now, let's check the correlation between features.

### Numeric features
Correlation matrix of numeric features:

```{r}
# Create the plot
numeric_correlation_matrix <- clinical_data |> 
  select(-code) |> 
  plot_corr_matrix(threshold = 0.5)

# Show the plot
numeric_correlation_matrix
```

### Categorical features
To address categorical features' colinearity, we will use **Cramer's V**.

```{r}
# Create the plot
categorical_correlation_matrix <- clinical_data |> 
  plot_corr_matrix(dtype = "cat", threshold = 0.5) |> 
  suppressWarnings()

# Show the plot
categorical_correlation_matrix
```

There is a perfect correlation between `heart_failure`-`LVEF` and a significant correlation between `hypolipidemic_meds`-`hypercholesterolemia`. Since the latter ones are more informative, we will drop `heart_failure` and `hypolipidemic_meds` feature.

```{r}
clinical_data <- clinical_data |> 
  select(-c(heart_failure, hypolipidemic_meds))
```

### Global correlation
To address global colinearity, we will use the **Variance Inflation Factor**:

$$VIF_i = \frac{1}{1 - R_i^2}$$

Where $R_i^2$ is the determination coefficient of the regression of the $i$-th feature on the remaining features.

```{r}
# Create plot 
VIF <- clinical_data |> 
    select(-c(code, chad2ds2_vasc_score, chads2_score, hatch_score)) |>
    plot_vif()

# Show the plot
VIF
```

Since $VIF < 5$ for all of the features, there is no multicollinearity.