---
title: "2.0- Exploratory Data Analysis"
output: html_document
---

# Configuration

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(skimr)
library(scales)
library(car)
```

# Load the data

```{r}
clinical_data <- read_parquet("data/intermediate/clinical_data.parquet")
head(clinical_data)
```

# Global summary
Let's check the column's data types and whether there are many missing values:

```{r}
skim(clinical_data) |> summary()
```

There are $720$ rows and $33$ columns: $21$ categorical (the target class can be found among these) and $12$ numerical features. Let's check each type separately.

## Numerical features
Let's sum up the data:

```{r}
clinical_data |> select(-code) |> skim() |> yank("numeric") |> 
  select(-c(starts_with("p"), hist, complete_rate))
```

Since most of the data in `LAD`, `LAA` and `LAV` is missing, we will drop those features:

```{r}
clinical_data <- clinical_data |> 
  select(-c(LAD, LAA, LAV))
```


## Categorical features

Let's sum up the data:

```{r}
skim(clinical_data) |> yank("factor") |> select(-c(complete_rate, ordered))
```

There are $17$ missing values in the `hypolipidemics` feature, $37$ in `left_atrial_enlarg` and $587$ in `af_trigger`. In this last case, since most of the data is missing, we will drop the feature:

```{r}
clinical_data <- clinical_data |> 
  select(-AF_trigger)
```


# Statistical analysis
Firstly, let's check how the data is distributed.

## Numerical features

```{r, fig.width=10, fig.height=8}
clinical_data |> select(-code) |> 
  select(where(is.numeric)) |> 
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value"
  ) |> 
  ggplot(aes(x = value)) +
  # 1. Histograma sutil de fondo (bajamos la opacidad a 0.4)
  # Usamos una paleta médica (ej. azul acero) en lugar del cyan puro que satura
  geom_histogram(
    aes(y = after_stat(density)), # Sincroniza la escala con la curva de densidad
    bins = 20,                    # Reducimos bins para evitar el efecto "peine desdentado"
    color = "white",              # Bordes blancos separadores más elegantes
    fill = "#2c3e50", 
    alpha = 0.4
  ) +
  # 2. Curva de densidad suavizada (elimina el problema visual de las barras sueltas)
  geom_density(
    color = "#16a085", 
    linewidth = 1, 
    fill = "#16a085", 
    alpha = 0.15
  ) +
  # 3. TRUCO CORRECCIÓN DECIMALES: labels = label_percent() o integers
  # Si el rango es pequeño, forzamos saltos enteros
  scale_x_continuous(
    labels = function(x) {
      # Eliminamos los valores NA para que no rompan la validación
      x_clean <- x[!is.na(x)]
      
      # Si todos los números del eje son enteros, quitamos decimales
      if (length(x_clean) > 0 && all(x_clean == round(x_clean))) {
        return(sprintf("%d", as.integer(x)))
      } else {
        # Si es continua real (ej. bmi), usamos el formato estándar de R
        return(labels(x)) 
      }
    }
  ) +
  facet_wrap(~ feature, scales = "free", ncol = 4) +
  theme_minimal(base_size = 11) +
  theme(
    # Limpieza de cuadrículas
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "#eaeded", linewidth = 0.4),
    
    # Consistencia con el diseño de tus boxplots (paneles estructurados)
    strip.background = element_rect(fill = "#f8f9f9", color = "#d5dbdb", linewidth = 0.5),
    strip.text = element_text(face = "bold", color = "#2c3e50", size = 10),
    
    # Línea base en el eje X para que descansen los histogramas
    axis.line.x = element_line(color = "#bdc3c7", linewidth = 0.6),
    axis.text = element_text(color = "#34495e"),
    axis.title = element_text(face = "bold", color = "#2c3e50")
  ) +
  labs(
    x = "",
    y = "",
    title = "Distribution of numeric features"
  )
```

Now, let's check how does the distribution of each feature behaves when we stratify the data by the target class:

```{r, fig.width=10, fig.height=8}
clinical_data |> select(-code) |> 
  select(any_of(c("AF_recurrence")), where(is.numeric)) |> 
  pivot_longer(
    cols = -AF_recurrence,
    names_to = "feature",
    values_to = "value"
  ) |> 
  ggplot(aes(x = AF_recurrence, y = value, fill = AF_recurrence)) +
  # Cajas con bordes definidos, transparencia y outliers sutiles
  geom_boxplot(
    alpha = 0.75, 
    color = "#2c3e50", 
    outlier.size = 1, 
    outlier.alpha = 0.4,
    linewidth = 0.6
  ) +
  # Escala logarítmica para manejar variables con rangos muy distintos
  scale_y_continuous(trans = pseudo_log_trans(base = 10)) +
  # Paleta de colores médica y profesional (puedes cambiar "D" por "A" o "E")
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.8) +
  # Cuadrícula modular con escalas libres
  facet_wrap(~ feature, scales = "free", ncol = 4) +
  # Basado en minimal pero súper vitaminado
  theme_minimal(base_size = 11) +
  theme(
    # Leyenda innecesaria porque el eje X ya dice la clase, la quitamos para ganar espacio
    legend.position = "none",
    
    # Limpiamos el fondo eliminando líneas secundarias pesadas
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "#eaeded", linewidth = 0.5),
    
    # Estilizamos las etiquetas de las facetas (los títulos de cada caja)
    strip.background = element_rect(fill = "#f8f9f9", color = "#d5dbdb", linewidth = 0.5),
    strip.text = element_text(face = "bold", color = "#2c3e50", size = 10),
    
    # Añadimos un sutil eje inferior para asentar los gráficos
    axis.line.x = element_line(color = "#bdc3c7", linewidth = 0.6),
    axis.text = element_text(color = "#34495e"),
    axis.title = element_text(face = "bold", color = "#2c3e50"),
    
    # Separación elegante entre las cajitas
    panel.spacing = unit(1.2, "lines")
  ) +
  labs(
    x = "",
    y = "",
    title = "Numeric features distributed by AF recurrence"
  )
```

## Categorical features

```{r, fig.width=16, fig.height=12}
clinical_data |> select(-code) |> 
  select(where(is.factor)) |> 
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value",
    values_transform = list(value = as.character)
  ) |> 
  mutate(value = fct_rev(fct_infreq(value))) |> 
  ggplot(aes(y = value, fill = value)) + 
  geom_bar(
    color = "#2c3e50", 
    alpha = 0.8, 
    width = 0.7,
    linewidth = 0.5
  ) +
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    hjust = -0.2,            
    size = 3.2,
    fontface = "bold",
    color = "#2c3e50"
  ) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "#f2f4f4", linewidth = 0.5),
    panel.grid.major.y = element_blank(),
    strip.background = element_rect(fill = "#f8f9f9", color = "#d5dbdb", linewidth = 0.5),
    strip.text = element_text(face = "bold", color = "#2c3e50", size = 10),
    axis.line.y = element_line(color = "#bdc3c7", linewidth = 0.6),
    axis.text = element_text(color = "#34495e", face = "bold"),
    axis.title = element_text(face = "bold", color = "#2c3e50"),
    panel.spacing = unit(1.5, "lines")
  ) +
  labs(
    x = "Counts",
    y = "",
    title = "Distribution of categorical features"
  )
```

Most of these categories are imbalanced, including the target class. Now let's see how this changes when we stratify the data by the target class:

```{r, fig.width=10, fig.height=16}
clinical_data |> 
  select(where(is.factor)) |> 
  pivot_longer(
    cols = -AF_recurrence,
    names_to = "feature",
    values_to = "value",
    values_transform = list(value = as.character)
  ) |> 
  filter(!is.na(value)) |> 
  # Ordenamos las categorías del eje X por su frecuencia total
  mutate(value = fct_infreq(value)) |> 
  
  ggplot(aes(x = value, fill = AF_recurrence)) +
  # Dibujamos las barras proporcionales al 100%
  geom_bar(
    color = "#2c3e50", 
    position = "fill",
    alpha = 0.85, 
    width = 0.7,
    linewidth = 0.5
  ) +
  # TRUCO INTERNO: Añadir el conteo absoluto de pacientes dentro de cada bloque de la barra
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    position = position_fill(vjust = 0.5), # Centra el número en medio de cada color
    size = 3.0,
    fontface = "bold",
    color = "white" # Blanco para que resalte sobre los colores oscuros de la barra
  ) +
  # Formateamos el eje Y para que muestre porcentajes reales (0% a 100%)
  scale_y_continuous(
    labels = label_percent(),
    expand = expansion(mult = c(0, 0.05)) # Margen mínimo porque el tope fijo es siempre 100%
  ) +
  # Paleta de colores consistente con tus boxplots (Viridis para la recidiva no/yes)
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.8) +
  
  # Liberamos el eje X para que cada panel acomode solo sus categorías reales
  facet_wrap(~ feature, scales = "free_x", ncol = 3) +
  
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top", # Colocamos la leyenda arriba para que no robe espacio a los lados
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(color = "#eaeded", linewidth = 0.5), 
    strip.background = element_rect(fill = "#f8f9f9", color = "#d5dbdb", linewidth = 0.5),
    strip.text = element_text(face = "bold", color = "#2c3e50", size = 10),
    axis.line.x = element_line(color = "#bdc3c7", linewidth = 0.6),
    axis.text.x = element_text(color = "#34495e", face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(color = "#34495e"),
    axis.title = element_text(face = "bold", color = "#2c3e50"),
    panel.spacing = unit(1.5, "lines")
  ) +
  labs(
    x = "",
    y = "",
    fill = "¿AF recurrence?",
    title = "Categorical features distributed by AF recurrence"
  )
```

## Multicolinearity
Finally, let's check the correlation between features to drop those that are highly correlated. To address numerical features' colinearity, we will use the **VIF**.

```{r}
# Convert the target class to numeric for the VIF calculation
temp_clinical_data <- clinical_data |> select(-code) |> 
  mutate(AF_recurrence = as.numeric(AF_recurrence))

# Fit a linear model with all the features
model <- lm(
    AF_recurrence ~ ., 
    data = temp_clinical_data
    )

# Calculate the equivalent VIF corrected for the degrees of freedom
vif_df <- vif(model) |> 
    as.data.frame() |> 
    rownames_to_column(var = "feature") |> 
    as_tibble() |> 
    rename(
        GVIF = 2,
        Df = 3,
        Adjusted_GVIF = 4
    ) |> 
    mutate(VIF_Equivalent = Adjusted_GVIF^2) |> 
    select(feature, VIF_Equivalent)

# Plot the VIF values
vif_df |> 
    ggplot(aes(x = reorder(feature, VIF_Equivalent), y = VIF_Equivalent)) + 
    geom_bar(stat = "identity", fill = "cyan", color = "black") +
    geom_hline(yintercept = 5, linetype = "dashed", color = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    labs(x = "", y = "VIF") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since a$VIF < 5$, there is no multicolinearity between the numeric features, so we will keep all of them.

# Missing values

# Save progress

```{r}
# Save the data
write_parquet(
    clinical_data, 
    "data/clean/clinical_data_clean.parquet"
    )

head(clinical_data)
```

