---
title: "3.0-Feature_Selection"
output: html_document
---

# Configuration

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(car)
```

# Load the data

```{r}
clinical_data <- read_parquet("data/intermediate/clinical_data_clean.parquet")
head(clinical_data)
```

# Feature selection
Given this data, we will analyze the statistical significance of the features to select those that are more relevant for the model. We will use binary logistic regression to check the significance of the features. We will keep those features with a p-value lower than 0.05.

Before running this analysis, let's drop the features that are not useful to train the model:

```{r}
clinical_data_selected_biv <- clinical_data |> 
  select(-c(fa_type))
```

## Bivariate model selection
Now, let's fit one logistic regression model and run the ANOVA test for each feature to check their statistical significance at predicting whether AF recurrence will take place:

```{r}
# Initialize vectors to store the features and their p-values
features <- clinical_data_selected_biv |> 
  select(-af_recurrence) |> 
  names()

p_values <- rep(0, length(features))

for (feature in features){
  if (feature == "af_recurrence") next else{
    
    # Create dynamic formula
      equation <- as.formula(paste("af_recurrence ~", feature))
            
    # Fit a logistic regression model with the current feature as the predictor
    model <- stats::glm(
        formula = equation, 
        data = clinical_data_selected_biv, 
        family = binomial
        )
    
    # Run ANOVA test to check the significance of the feature
    anova_result <- anova(model, test = "Chisq")
    
    # Extract the p-value of the feature
    p_values[which(features == feature)] <- anova_result[2, 5]
    }
  }

# Create a data frame with the features and their p-values
binary_logist_model <- data.frame(
  feature = features,
  p_value = p_values
  ) |>
  arrange(p_value)

binary_logist_model
```

Given a determined p-value threshold, we can select the features that are statistically significant for predicting the target class. In this case, we will keep those features with a p-value lower than:

```{r}
# Set the p-value threshold for feature selection
threshold <- 0.05

# Select features with p-value lower than the threshold
features_to_keep <- binary_logist_model |> 
  filter(p_value < threshold) |> 
  pull(feature)

# Show the selected features and their p-values
binary_logist_model |> filter(p_value < threshold)
```

## Multivariate model selection
Finally, let's fit a multivariate logistic regression model with the features that have a p-value lower than the threshold and check their significance again:

```{r}
# Add target for the multivariate logistic regression model
clinical_data_selected_multiv <- clinical_data_selected_biv |> 
  select(all_of(features_to_keep), af_recurrence)

# Fit a multivariate logistic regression model with the selected features
multiv_logist_model <- glm(
  af_recurrence ~ .,
  data = clinical_data_selected_multiv,
  family = binomial
  )

# Run ANOVA test and extract the p-values of the features
p_values_multiv <- Anova(multiv_logist_model, type = "II") |> 
  pull("Pr(>Chisq)")

# Create a data frame with the results
multiv_logist_model <- data.frame(
  feature = features_to_keep,
  p_value = p_values_multiv
  ) |>
  arrange(p_value)

multiv_logist_model
```

Let's drop the features that are not statistically significant in the multivariate model:

```{r}
# Set the p-value threshold for feature selection
threshold2 <- 0.2

# Select features with p-value lower than the threshold
ultimate_features <- multiv_logist_model |> 
  filter(p_value < threshold2) |> 
  pull(feature)

# Show the selected features and their p-values
multiv_logist_model |> filter(p_value < threshold)
```

# Final data set
Finally, let's save the data set with the selected features:

```{r}
clinical_data_ultimate <- clinical_data_selected_multiv |> 
  select(all_of(ultimate_features), af_recurrence)

clinical_data_ultimate |> 
  write_parquet("data/processed/clinical_data_selected_features.parquet")

head(clinical_data_ultimate)
```
